---
output: github_document
Date: 20/08/2021
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rddharmony

The goal of rddharmony is to implement a workflow for births, deaths and population data extracted from vital registration databases and census. This workflow includes extracting data from DemoData, harmonizing age groups, identifying full series, validating totals and by sex, eventually producing clean and harmonised datasets for each location and each data type (births, deaths or population).


## Installation

You can install the released version of rddharmony from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("rddharmony")
```
_Note: This package is not yet on CRAN_

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Shelmith-Kariuki/rddharmony")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(rddharmony)
## To insert example later
```
