---
title: "Validating totals over sex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating totals over sex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

The purpose of this process is to check if counts by sex is equal to counts for both sexes and adjust in case of any difference i.e if Female counts + Male counts is different from the population for Both sexes, then Both sexes = Female + Male

There are two functions involved. `dd_validate_totals_over_sex` is implemented on births and deaths data while `dd_validate_totals_over_sex_new` is implemented on population data.

The functions are defined as follows:

> dd_validate_totals_over_sex(data) [^1]

> dd_validate_totals_over_sex_new(data) [^2]

## Part 1: Validating totals over sex: births and deaths

To show how `dd_validate_totals_over_sex_bd` works, we will be using the `validate_totals_over_sex` dataset that is part of this package. This dataset contains death counts of Kuwait for the year 2009, obtained from the Demographic Yearbook. 

```{r setup}
## Load the packages required
library(rddharmony)
library(kableExtra)
library(dplyr)
library(purrr)
library(tidyr)

## Create a function to be used to generate the table output
tab_output <- function(tab) {
  kable(tab, booktabs = TRUE, align = "c", table.envir = "capctable", longtable = TRUE, row.names = FALSE) %>%
    kable_styling() %>%
    row_spec(0, bold = T, color = "white", background = "#6ebed8") %>%
    kable_paper(html_font = "helvetica") %>%
    scroll_box(width = "100%", height = "300px")
}
```

```{r}
df <- validate_totals_over_sex_bd %>% 
  select(abridged, complete, SexID, AgeStart, AgeEnd, AgeSpan,AgeSort, AgeLabel,DataValue)

df %>% tab_output()
```

From the data, it can be seen that that `Male` counts and `Female` counts add up to `Both sexes` counts. For demonstration purposes, we will alter the data so that it can represent an imperfect situation where the totals do not add up. We will edit the abridged `Both sexes` Total value from  6266 to 8978 and the complete `Female` Total value from 2460 to 3810.

```{r}
df <- df %>% 
      mutate(DataValue = ifelse(abridged == TRUE & AgeLabel == "Total" & SexID == 3, 8978, DataValue),
             DataValue = ifelse(complete == TRUE & AgeLabel == "Total" & SexID == 2, 3810, DataValue))
df %>% tab_output()
```

This dataset contains both abridged and complete series, so we will begin with the abridged series. 

```{r}
df_abr <- df %>% 
  filter(abridged == TRUE)
df_abr %>% tab_output()
```

We begin by checking if data for both males and females exists, and whether both series are full [^3]

```{r}
    full_m <- dd_series_isfull(data = df_abr[df_abr$SexID == 1,], abridged = TRUE)
    full_f <- dd_series_isfull(data = df_abr[df_abr$SexID == 2,], abridged = TRUE)

```

Next, check whether all the age labels in the males data are present in the females data.

```{r}
    ages_same <- all(df_abr$AgeLabel[df_abr$SexID == 1] %in% df_abr$AgeLabel[df_abr$SexID == 2])

```

If data for both males and females exists and are full, and all the age labels in the males data are present in the females data,
re-calculate the values to ensure that `Both sexes` is the sum of `Female` and `Male` data. 

```{r}
    if (all(full_m, full_f, ages_same)) {
    df_abr <- df_abr %>%
      mutate(SexID = paste0("X", SexID)) %>%
      spread(key = SexID, value = DataValue) %>%
      mutate(X3 = X1 + X2) %>% # ensure that both sexes is sum of male and female
      gather(key = "SexID", value = "DataValue", c("X1","X2","X3")) %>%
      mutate(SexID = as.numeric(substr(SexID,2,2)),
             note = NA) %>%
      dplyr::filter(!is.na(DataValue)) %>% # need this bc NAs are introduced if both sexes has more age groups than by sex
      arrange(SexID, AgeSort)
    } else {
      df_abr <- df_abr %>%
        mutate(note = "Inconsistent age groups for males and females; Both sexes not reconciled")
    }
df_abr %>% tab_output()
```

We now have the correct value for `Both sexes` total. 

```{r}
df_abr$DataValue[df_abr$abridged == TRUE & df_abr$AgeLabel == "Total" & df_abr$SexID == 3]
```
The same process is repeated for the complete series

```{r}
df_cpl <- df %>% 
  filter(complete == TRUE)
df_cpl %>% tab_output()

```

Check that data for both males and females exists, and that both series are full 

```{r}
    full_m <- dd_series_isfull(data = df_cpl[df_cpl$SexID == 1,], abridged = TRUE)
    full_f <- dd_series_isfull(data = df_cpl[df_cpl$SexID == 2,], abridged = TRUE)

```

Check that all the age labels in the males data are present in the females data.

```{r}
    ages_same <- all(df_cpl$AgeLabel[df_cpl$SexID == 1] %in% df_cpl$AgeLabel[df_cpl$SexID == 2])

```

And if the above condition are met, re-calculate the values to ensure that that they add up.

```{r}
    if (all(full_m, full_f, ages_same)) {
    df_cpl <- df_cpl %>%
      mutate(SexID = paste0("X", SexID)) %>%
      spread(key = SexID, value = DataValue) %>%
      mutate(X3 = X1 + X2) %>% # ensure that both sexes is sum of male and female
      gather(key = "SexID", value = "DataValue", c("X1","X2","X3")) %>%
      mutate(SexID = as.numeric(substr(SexID,2,2)),
             note = NA) %>%
      dplyr::filter(!is.na(DataValue)) %>% # need this bc NAs are introduced if both sexes has more age groups than by sex
      arrange(SexID, AgeSort)
    }
df_cpl %>% tab_output()

```

In cases where all age labels in female data are not present in male data or both series are not full, a note is generated alerting the user of the same issue. 

```{r}
# df_cpl <- df_cpl %>%
#         mutate(note = "Inconsistent age groups for males and females; Both sexes not reconciled")
```

## Part 2: Validating totals over sex: population counts

To demonstrate this process for population counts, we will make use of the `validate_totals_over_sex_pop` dataset. This data represents Myanmar population counts for the year 2014 extracted from census records.

```{r}
df <- validate_totals_over_sex_pop %>% 
  select(abridged, complete, SexID, AgeStart, AgeEnd, AgeSpan,AgeSort, AgeLabel,DataValue)

df %>% tab_output()
```


The process here is a bit different in that we first harmonize open age groups by truncating to minimum open age group of males and females and later validate the totals by sex.

We begin by subsetting the data to only have abridged records. 

```{r}
  abr <- validate_totals_over_sex_pop %>%
    dplyr::filter(abridged == TRUE)

```

We then determine the minimum open age group in the males and females data and extract the record containing this minimum age.

```{r}
  oa_min <- suppressWarnings(min(abr$AgeStart[abr$AgeSpan == -1 & abr$AgeLabel != "Total" & abr$SexID %in% c(1,2)]))
  oa_record <- abr %>%
    dplyr::filter(AgeStart == oa_min & AgeSpan == -1) %>%
    select(-SexID, -DataValue) %>%
    distinct()
```

We go ahead and calculate the summation of data values of age labels that exceed this minimum open age group for each of the sex ids and store the result in different datasets.

```{r}
if (1 %in% abr$SexID) {
    oa_record_m <- oa_record %>%
      mutate(SexID =1,
             DataValue = sum(abr$DataValue[abr$AgeStart >= oa_min & abr$SexID ==1]))
  } else {
    oa_record_m <- NULL
  }
  if (2 %in% abr$SexID) {
    oa_record_f <- oa_record %>%
      mutate(SexID =2,
             DataValue = sum(abr$DataValue[abr$AgeStart >= oa_min & abr$SexID ==2]))
  } else {
    oa_record_f <- NULL
  }
  if (3 %in% abr$SexID) {
    oa_record_b <- oa_record %>%
      mutate(SexID =3,
             DataValue = sum(abr$DataValue[abr$AgeStart >= oa_min & abr$SexID ==3]))
  } else {
    oa_record_b <- NULL
  }

  if (0 %in% abr$SexID) {
    oa_record_o <- oa_record %>%
      mutate(SexID =0,
             DataValue = sum(abr$DataValue[abr$AgeStart >= oa_min & abr$SexID ==0]))
  } else {
    oa_record_o <- NULL
  }
```

Later, we drop all the age labels that exceed this minimum open age group, and append the resulting data with the datasets generated above.

```{r}
  abr <- abr %>%
    dplyr::filter(AgeStart < oa_min) %>%
    dplyr::filter(!(AgeSpan == -1 & AgeLabel != "Total" & AgeStart < oa_min)) %>% # remove any oa that is below oa_min 
    bind_rows(oa_record_m, oa_record_f, oa_record_b, oa_record_o) %>%
    arrange(SexID, AgeStart)
```

This process is repeated for the complete series ... 

```{r}
cpl <- validate_totals_over_sex_pop %>%
    dplyr::filter(complete == TRUE)

  if (nrow(cpl) > 50){ # only bother with this if complete series is usable

      oa_min <- suppressWarnings(min(cpl$AgeStart[cpl$AgeSpan == -1 & cpl$AgeLabel != "Total" & cpl$SexID %in% c(1,2)]))
      
      oa_record <- cpl %>%
        dplyr::filter(AgeStart == oa_min & AgeSpan == -1) %>%
        select(-SexID, -DataValue) %>%
        distinct()
      
      if (1 %in% cpl$SexID) {
      oa_record_m <- oa_record %>%
        mutate(SexID =1,
               DataValue = sum(cpl$DataValue[cpl$AgeStart >= oa_min & cpl$SexID ==1]))
      } else {
        oa_record_m <- NULL
      }
      if (2 %in% cpl$SexID) {
      oa_record_f <- oa_record %>%
        mutate(SexID =2,
               DataValue = sum(cpl$DataValue[cpl$AgeStart >= oa_min & cpl$SexID ==2]))
      } else {
        oa_record_f <- NULL
      }
      if (3 %in% cpl$SexID) {
      oa_record_b <- oa_record %>%
        mutate(SexID =3,
               DataValue = sum(cpl$DataValue[cpl$AgeStart >= oa_min & cpl$SexID ==3]))
      } else {
        oa_record_b <- NULL
      }
      if (0 %in% cpl$SexID) {
      oa_record_o <- oa_record %>%
        mutate(SexID =0,
               DataValue = sum(cpl$DataValue[cpl$AgeStart >= oa_min & cpl$SexID ==0]))
      } else {
        oa_record_o <- NULL
      }

      cpl <- cpl %>%
        dplyr::filter(AgeStart < oa_min) %>%
        dplyr::filter(!(AgeSpan == -1 & AgeLabel != "Total" & AgeStart < oa_min)) %>%
        bind_rows(oa_record_m, oa_record_f, oa_record_b, oa_record_o) %>%
        arrange(SexID, AgeStart)
  } else {
    cpl <- NULL
  }

```

...  and the two resulting datasets are appended to form one.

```{r}
  df <- rbind(abr, cpl)

```

It is at this point that the validation of totals by sex takes place. 

```{r}
# now validate totals over sex
  df <- df %>%
    mutate(SexID = paste0("X", SexID)) %>%
    spread(key = SexID, value = DataValue)

  if (!("X0" %in% names(df))) {
    df$X0 <- 0
  }
  if (!("X1" %in% names(df))) {
    df$X1 <- 0
  }
  if (!("X2" %in% names(df))) {
    df$X2 <- 0
  }
  if (!("X3" %in% names(df))) {
    df$X3 <- 0
  }


  data.out <- df %>%
    mutate(X0 = replace(X0, is.na(X0), 0),
           X1 = replace(X1, is.na(X1), 0),
           X2 = replace(X2, is.na(X2), 0),
           X3 = replace(X3, is.na(X3), 0),
           total_over_sex = X0 + X1 + X2,
           X3 = ifelse(X3 > total_over_sex, X3, total_over_sex), # total as max of reported total or sum over sex
           X1 = ifelse(X1 == 0 & total_over_sex > 0 & X3 > total_over_sex, X3 - total_over_sex, X1),
           total_over_sex = X0 + X1 + X2,
           X2 = ifelse(X2 == 0 & total_over_sex > 0 & X3 > total_over_sex, X3 - total_over_sex, X2),
           total_over_sex = X0 + X1 + X2,
           X0 = ifelse(X3 > total_over_sex & total_over_sex > 0, X0 + (X3 - total_over_sex), X0 ),
           pct_m = X1 / (X1 + X2), # percentage of males among males plus females
           pct_m = replace(pct_m, is.na(pct_m), 0), # avoids divide by zero problems
           X1 = X1 + (X0 * pct_m),
           X2 = X2 + (X0 * (1-pct_m)),
           X3 = X1 + X2) %>%
    select(-X0, -total_over_sex, -pct_m) %>%
    gather(key = "SexID", value = "DataValue", c("X1","X2","X3")) %>%
    mutate(SexID = as.numeric(substr(SexID,2,2)),
           note = NA) %>%
    arrange(SexID, AgeSort)

```


[^1]: [dd_validate_totals_over_sex](https://shelmith-kariuki.github.io/rddharmony/reference/dd_validate_totals_over_sex.html)

[^2]: [dd_validate_totals_over_sex_new](https://shelmith-kariuki.github.io/rddharmony/reference/dd_validate_totals_over_sex_new.html)

[^3]: [dd_series_isfull](https://shelmith-kariuki.github.io/rddharmony/reference/dd_series_isfull.html)
