---
title: "Validating totals over age"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating totals over age}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

The purpose of this function is to reconcile `Totals`, in cases where reported values are different from calculated values, after having standardized age series. If the computed total is greater than reported total, the reported total is replaced with the computed total. If computed is less than reported, the difference is added to "Unknown" age

This function is defined as follows:

> dd_validate_totals_over_age(data) [^1]

To show how it works, we shall use the `vitals_std_full_sample` dataset that is embedded on this package. This dataset represents a scenario where we have both abridged and complete series for the same unique id. 

```{r setup}
## Load the packages required
library(rddharmony)
library(kableExtra)
library(dplyr)
library(purrr)

## Create a function to be used to generate the table output
tab_output <- function(tab) {
  kable(tab, booktabs = TRUE, align = "c", table.envir = "capctable", longtable = TRUE, row.names = FALSE) %>%
    kable_styling() %>%
    row_spec(0, bold = T, color = "white", background = "#6ebed8") %>%
    kable_paper(html_font = "helvetica") %>%
    scroll_box(width = "100%", height = "300px")
}
```

```{r}
data <- vitals_std_full_sample %>% 
       select(starts_with("Age"), SexID, DataValue, abridged, complete)
data %>% tab_output()
```

This process begins by subsetting the data to have abridged records for each particular sex id. The reported totals are extracted and actual totals calculated. If the reported and actual totals exist and if the computed total is greater than reported total, then the reported total is replaced with the computed. If the computed total is less than the reported, then the difference is added to the "Unknown" age label. 

```{r}
sex <- 3 

    ## subset the data to have abridged records for each particular sex id
    df_abr <- data %>%
      dplyr::filter(SexID == sex & abridged == TRUE)

    ## extract the reported totals and the actual totals
    total_reported_abr <- df_abr$DataValue[df_abr$AgeLabel == "Total"]
    total_computed_abr <- sum(df_abr$DataValue[df_abr$AgeSpan %in% c(-1,-2, 5) & df_abr$AgeLabel != "Total"])


    ## if the reported and actual totals exist...
    if (!is_empty(total_reported_abr) & !is_empty(total_computed_abr)) {

      # if computed is greater than reported, then replace reported with computed
      diff <- total_computed_abr - total_reported_abr
      if (diff > 0 ) {
        df_abr$DataValue[df_abr$AgeLabel == "Total"] <- total_computed_abr
      }
      # if computed is less than reported, then add difference to "Unknown" age
      if (diff < 0 ) {
        df_abr$DataValue[df_abr$AgeLabel == "Unknown"] <- df_abr$DataValue[df_abr$AgeLabel == "Unknown"] +
          (total_reported_abr-total_computed_abr)
      }

    }
```

The same process is repeated for complete records

```{r}
    ## subset the data to have complete records for each particular sex id
    df_cpl <- data %>%
      dplyr::filter(SexID == sex & complete == TRUE)

    total_reported_cpl <- df_cpl$DataValue[df_cpl$AgeLabel == "Total"]
    total_computed_cpl <- sum(df_cpl$DataValue[df_cpl$AgeLabel != "Total"])

    if (!is_empty(total_reported_cpl) & !is_empty(total_computed_cpl)) {

      # if computed is greater than reported, then replace reported with computed
      diff <- total_computed_cpl - total_reported_cpl
      if (diff > 0 ) {
        df_cpl$DataValue[df_cpl$AgeLabel == "Total"] <- total_computed_cpl
      }
      # if computed is less than reported, then add difference to "Unknown" age
      if (diff < 0 ) {
        df_cpl$DataValue[df_cpl$AgeLabel == "Unknown"] <- df_cpl$DataValue[df_cpl$AgeLabel == "Unknown"] +
          (total_reported_cpl-total_computed_cpl)
      }

    }
```

Eventually, both datasets are appended to form the final output

```{r}
out.data <- NULL
out.data <- rbind(out.data, df_abr, df_cpl)
out.data %>% tab_output()
```

[^1]: [dd_validate_totals_over_age](https://shelmith-kariuki.github.io/rddharmony/reference/dd_validate_totals_over_age.html)
