---
title: "Reconciling abridged and complete series"
output: rmarkdown::html_vignette
description: >
  This function reconciles abridged and complete series. 
vignette: >
  %\VignetteIndexEntry{Reconciling abridged and complete series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

This function is implemented in cases where we have both abridged and complete series. It is defined as follows:

> DDharmonize_AbridgedAndComplete(data_abr, data_cpl_from_abr, data_cpl) [^1]

To show how it works, we shall use the `data_abr` and `data_cpl` datasets that are embedded on this package. For this vignette, we shall assume that `data_cpl_from_abr == NULL` i.e we do not have complete records obtained from abridged series. This is very common in the deaths data. 

```{r setup}
## Load the packages required
library(rddharmony)
library(kableExtra)
library(dplyr)
library(purrr)

## Create a function to be used to generate the table output
tab_output <- function(tab) {
  kable(tab, booktabs = TRUE, align = "c", table.envir = "capctable", longtable = TRUE, row.names = FALSE) %>%
    kable_styling() %>%
    row_spec(0, bold = T, color = "white", background = "#6ebed8") %>%
    kable_paper(html_font = "helvetica") %>%
    scroll_box(width = "100%", height = "300px")
}
```

The abridged data consists of 5-year age groups ranging from "10-14" to "50+" with a total value of 114870 

```{r}
data_abr %>%
  select(-note) %>%
  tab_output()
```

... and the complete series consists of single years of age ranging from "10" to "50+" with a total value of 114870 as well.

```{r}
data_cpl %>%
  select(-note) %>%
  tab_output()

```

As noted above, we assume that we do not have complete records obtained from abridged series, so we set this to NULL.

```{r}
data_cpl_from_abr <- NULL
```


This function should be looped over each of the Sex Ids but in this vignette, we will only be handling SexId == 3 (Both sexes). 

```{r}
sex <- 3
abr_sex <- NULL ## will contain the final abridged records
cpl_sex <- NULL ## will contain the final complete records
```

The first step in this function involves creating flags to check whether each of the datasets listed in the function arguments is available. 

```{r}

has_abr <- nrow(data_abr[!is.na(data_abr$DataValue) & data_abr$SexID == sex, ]) > 0
has_abr <- ifelse(is_empty(has_abr), FALSE, has_abr)
has_cpl_from_abr <- nrow(data_cpl_from_abr[!is.na(data_cpl_from_abr$DataValue) & data_cpl_from_abr$SexID == sex, ]) > 0
has_cpl_from_abr <- ifelse(is_empty(has_cpl_from_abr), FALSE, has_cpl_from_abr)
has_cpl <- nrow(data_cpl[!is.na(data_cpl$DataValue) & data_cpl$SexID == sex, ]) > 0
has_cpl <- ifelse(is_empty(has_cpl), FALSE, has_cpl)

cat("has_abr: ", has_abr, "\n")
cat("has_cpl: ", has_cpl, "\n")
cat("has_cpl_from_abr: ", has_cpl_from_abr, "\n")
```

If the abridged series exists, extract the data value of AgeLabel == "Total" for each of the Sex Ids.

```{r}
if (has_abr) {
  df_abr <- data_abr %>%
    dplyr::filter(SexID == sex) %>%
    select(-note, -SexID)
  total_abr <- df_abr$DataValue[df_abr$AgeLabel == "Total"]
} else {
  df_abr <- NULL
}
total_abr
```

If complete records obtained from abridged series exist, extract them for each of the Sex Ids. We know that this does not exist in our case so `df_cpl_from_abr` will be NULL.

```{r}
if (has_cpl_from_abr) {
  df_cpl_from_abr <- data_cpl_from_abr %>%
    dplyr::filter(SexID == sex) %>%
    select(-note, -SexID)
} else {
  df_cpl_from_abr <- NULL
}
```

If complete series exists, subset the data to only have records for the specific sex being looped over and extract the data value of AgeLabel == "Total". If the complete series doesn't exist but complete records obtained from abridged dataset does `df_cpl_from_abr`, set the former to the latter and generate abridged records from these complete records [^2]. 

```{r}
if (has_cpl) {
  df_cpl <- data_cpl %>%
    dplyr::filter(SexID == sex) %>%
    select(-note, -SexID)
  total_cpl <- df_cpl$DataValue[df_cpl$AgeLabel == "Total"]
  df_abr_from_cpl <- df_cpl %>% dd_single2abridged()
} else { # if no data_cpl, df_cpl will be equal to df_cpl_from_abr (complete records obtained from abridged)
  if (has_cpl_from_abr) {
    # use data_cpl_from_abr
    if (all(data_cpl_from_abr$AgeSpan < 0)) {
      data_cpl_from_abr <- NULL
    }
    df_cpl <- df_cpl_from_abr
    if (!is.null(df_cpl)) {
      total_cpl <- df_cpl$DataValue[df_cpl$AgeLabel == "Total"]
      df_abr_from_cpl <- df_cpl %>% dd_single2abridged()
      has_cpl <- TRUE
    }
  } else {
    df_cpl <- NULL
    df_abr_from_cpl <- NULL
  }
}
```

Next, where both abridged and complete series exist and are not empty, we check if the totals match.

```{r}
if (has_abr & has_cpl) {
  total_diff <- total_abr - total_cpl
  # total_match <- total_diff == 0
  total_match <- total_diff <= 0.5
  if (is_empty(total_match)) {
    total_match <- TRUE
  }
} else {
  total_match <- FALSE
}

if (exists("total_abr") & exists("total_cpl")) {
  total_match <- ifelse(is_empty(total_abr) & is_empty(total_cpl), FALSE, total_match)
}
total_match
```

If they do, we proceed to reconcile abridged series with records from complete. This step involves first appending the abridged series with the abridged records derived from complete series, but only for the age labels that do not exist in the former, to avoid duplication. Open age groups that do not close the series are dropped and all possible open age groups
are computed [^4]. We again drop records for open age groups that do not close the series, check to see whether the series is full [^5] and eventually generate a note that alerts the user that the series is missing some data when the series is not full.

```{r}
if (total_match) {
  # append the abridged series with the abridged records derived from complete series, but only for the
  # age labels that do not exist in the former, to avoid duplication
  df_abr <- df_abr %>%
    bind_rows(df_abr_from_cpl %>% dplyr::filter(!(AgeLabel %in% df_abr$AgeLabel)))

  # drop records for open age groups that do not close the series
  oag_start_abr <- dd_oag_agestart(df_abr, multiple5 = TRUE)

  if (!is_empty(oag_start_abr)) {
    df_abr <- df_abr %>%
      dplyr::filter(!(AgeStart > 0 & AgeSpan == -1 & AgeStart != oag_start_abr))
  }

  # compute all possible open age groups
  oag_abr <- dd_oag_compute(df_abr, age_span = 5)
  if (!is.null(oag_abr)) {
    df_abr <- df_abr %>%
      bind_rows(oag_abr %>% dplyr::filter(!(AgeLabel %in% df_abr$AgeLabel)))
  }

  # drop records for open age groups that do not close the series
  oag_start_abr <- dd_oag_agestart(df_abr, multiple5 = TRUE)
  if (!is_empty(oag_start_abr)) {
    df_abr <- df_abr %>%
      dplyr::filter(!(AgeStart > 0 & AgeSpan == -1 & AgeStart != oag_start_abr)) %>%
      mutate(series = "abridged reconciled with complete") %>%
      arrange(AgeSort)
  }

  # check to see whether the series is full
  isfull_abr <- dd_series_isfull(df_abr, abridged = TRUE)

  df_abr$note <- ifelse(isfull_abr, NA, "The abridged series is missing data for one or more age groups.")
  df_abr$SexID <- sex
}
```

Once that is done, we go ahead and reconcile complete series with records from abridged. We start by appending the complete series with the complete records derived from abridged series, but only for the age labels that do not exist in the former, to avoid duplication. Open age groups that do not close the series are dropped , all possible open age groups computed and an open age group that is a multiple of five is computed. A note is also generated alerting the user of any instances of missing data. If the only remaining record on complete is "Unknown" or "Total" , the whole series is discarded.

```{r}
if (total_match) {
  if (!is.null(df_cpl_from_abr)) {
    df_cpl <- df_cpl %>%
      bind_rows(df_cpl_from_abr %>% dplyr::filter(!(AgeLabel %in% df_cpl$AgeLabel)))
  }

  # only process if there are multiple closed age groups in the series
  if (nrow(df_cpl[df_cpl$AgeSpan == 1, ]) > 1) {

    # drop records for open age groups that do not close the series
    oag_start_cpl <- dd_oag_agestart(df_cpl, multiple5 = FALSE)
    df_cpl <- df_cpl %>%
      dplyr::filter(!(AgeStart > 0 & AgeSpan == -1 & AgeStart != oag_start_cpl))

    # compute all possible open age groups
    oag_cpl <- dd_oag_compute(df_cpl, age_span = 1)
    if (!is.null(oag_cpl)) {
      df_cpl <- df_cpl %>%
        bind_rows(oag_cpl %>% dplyr::filter(!(AgeLabel %in% df_cpl$AgeLabel)))
    }

    # identify the open age group that is a multiple of five
    oag_start_cpl <- dd_oag_agestart(df_cpl, multiple5 = TRUE)

    df_cpl <- df_cpl %>%
      dplyr::filter(!(AgeStart > 0 & AgeSpan == -1 & AgeStart != oag_start_cpl)) %>%
      dplyr::filter(!(AgeSpan == 1 & AgeStart >= oag_start_cpl)) %>%
      mutate(series = "complete reconciled with abridged") %>%
      arrange(AgeSort)

    isfull_cpl <- dd_series_isfull(df_cpl, abridged = FALSE)

    df_cpl$note <- ifelse(isfull_cpl, NA, "The complete series is missing data for one or more age groups.")
    df_cpl$SexID <- sex
  } else {
    df_cpl <- NULL
  }

  # if the only remaining record on complete is "Unknown" or "Total" then discard the whole series

  if (all(unique(df_cpl$AgeLabel) %in% c("Unknown", "Total"))) {
    df_cpl <- NULL
  }
}
```

In a case where totals do not match, a note is generated alerting the user about the mismatch.

```{r}

if (!total_match) {

if (!is.null(df_abr)) {
  df_abr$note <- "Different totals on abridged and complete preclude reconciliation"
  df_abr$SexID <- sex
}
if (!is.null(df_cpl)) {
  df_cpl$note <- "Different totals on abridged and complete preclude reconciliation"
  df_cpl$SexID <- sex
}
}
```

This process is repeated for each of the Sex Ids and abridged and complete variables are generated depending on the type of data.

```{r}
abr_sex <- rbind(abr_sex, df_abr)
cpl_sex <- rbind(cpl_sex, df_cpl)

## generate abridged and complete variables which are TRUE/FALSE depending on the series
if (!is.null(abr_sex)) {
  abr_sex <- abr_sex %>%
    mutate(
      abridged = TRUE,
      complete = FALSE
    )
}
if (!is.null(cpl_sex)) {
  cpl_sex <- cpl_sex %>%
    mutate(
      abridged = FALSE,
      complete = TRUE
    )
}
```

Eventually, both data sets (abridged records and complete records) are appended to form one master dataset

```{r}
outdata <- rbind(
  abr_sex,
  cpl_sex
)

outdata %>% tab_output()
```

[^1]: [DDharmonize_AbridgedAndComplete](https://shelmith-kariuki.github.io/rddharmony/reference/DDharmonize_AbridgedAndComplete.html)

[^2]: [dd_single2abridged](https://shelmith-kariuki.github.io/rddharmony/reference/dd_single2abridged.html)

[^3]: [dd_oag_agestart](https://shelmith-kariuki.github.io/rddharmony/reference/dd_oag_agestart.html)

[^4]: [dd_oag_compute](https://shelmith-kariuki.github.io/rddharmony/articles/dd_oag_compute.html)

[^5]: [dd_series_isfull](https://shelmith-kariuki.github.io/rddharmony/reference/dd_series_isfull.html)
